### 2023-09-01 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_betacov_mapping@6e36402c1b5194bfa5ebc5948ea574144eaefacd 🚀

We downloaded the data on bats hosts of \emph{Betacoronavirus} from
\texttt{https://www.viralemergence.org/betacov} on
Apr.~2022,\textsuperscript{\textbf{Becker2022OptPre?}} and filtered it
to ``known'' hosts (established before the emergence of SARS-CoV-2) and
``novel'' hosts (confirmed through sampling and competence assays since
the initial data collection). The original database was assembled by a
combination of data mining and literature surveys, including automated
alerts on the ``bats'' and ``coronavirus'' keywords to identify novel
empirical evidence of bats-betacoronaviruses associations; this yielded
a total of 126 known hosts, 47 of which were novel hosts. This
host--virus list of interactions was obtained through a comprehensive
aggregation of GenBank data as well as systematic literature
searches,\textsuperscript{\textbf{Becker2022OptPre?},\textbf{Cohen2022SamStr?}}
such that we have high confidence in its fitness for the purpose of
inference at a large spatial scale.

@@

We downloaded the rangemap of every current bat species that was
classified as an empirically documented host of \emph{Betacoronavirus}
from the previous step, according to recent IUCN
data.\textsuperscript{\textbf{IUCN2021IucRed?}} The IUCN data have been
assembled to support wildlife conservation efforts, and therefore we do
not expect that they are biased by wildlife disease sampling efforts or
priority. The range maps were subsequently rasterized using the
\texttt{rasterize} function from
\texttt{GDAL}\textsuperscript{\textbf{RouaultEven2022GdaOgr?}} at a
resolution of approximately 100kmx100km at the equator. For every pixel
in the resulting raster where at least one bat host of
\emph{Betacoronavirus} was present, we extract the species pool (list of
all known bat hosts), which was used to calculate the following risk
assessment components: bat phylogenetic diversity, bat compositional
uniqueness, and predicted viral sharing risk.

@@

For every pixel, we measured Faith's Phylogenetic
Diversity\textsuperscript{\textbf{Faith1992ConEva?}} based on a recent
synthetic tree with robust time calibration, covering about 6000
mammalian species.\textsuperscript{\textbf{Upham2019InfMam?}} Faith's PD
measures the sum of unique branches from an arbitrary root to a set of
tips, and comparatively larger values indicate a more phylogenetic
diverse species pool. We measured phylogenetic diversity starting from
the root of the entire tree (and not from Chiroptera); this bears no
consequences on the resulting values, since all branches leading up to
Chiroptera are only counted once per species pool, and (as we explain
when describing the assembly of the composite risk map), all individual
risk components are ranged in {[}0,1{]}. This measure incorporates a
richness component, which we chose not to correct for; the
interpretation of the phylogenetic diversity is therefore a weighted
species richness, that accounts for phylogenetic over/under-dispersal in
some places.

@@

For every species pool, we measured its Local Contribution to
Beta-Diversity;\textsuperscript{\textbf{Legendre2013BetDiv?}} LCBD works
from a species-data matrix (traditionally noted as \(\mathbf{Y}\)),
where species are rows and sites are columns, and a value of 1 indicates
occurrence. We extracted the Y matrix assuming that every pixel
represents a unique location, and following best
practices\textsuperscript{\textbf{Legendre2019SpaTem?}} transformed it
using Hellinger's distance to account for unequal bat richness at
different pixels. The correction of raw community data is particularly
important for two reasons: first, it prevents the artifact of richer
sites having higher importance; second, it removes the effect of overall
species richness, which is already incorporated in the phylogenetic
diversity component. High values of LCBD indicate that the pixel has a
community that is on average more dissimilar in species composition than
what is expected knowing the entire matrix, i.e.~a more unique
community. Recent results
by\textsuperscript{\textbf{Dansereau2022EvaEco?}} shows that LCBD
measures are robust with regards to spatial scale, and are therefore
applicable at the global scale.

@@

For all bat hosts of \emph{Betacoronavirus}, we extracted their
predicted viral sharing network, generated from a previously published
generalized additive mixed model of virus sharing by a tensor function
of phylogenetic distance and geographic range overlap across
mammals.\textsuperscript{\textbf{Albery2020PreGlo?}} This network stores
pairwise values of viral community similarity, measured for all hosts
(to maintain consistency with teh phylogenetic diversity measure) across
all viruses; therefore, we consider that it accounts for some overall
similarity in the way hosts deal with viruses, and not only
betacoronaviruses. There is empirical evidence that capacity for
cross-species transmission even between divergent species is generally
high,\textsuperscript{55} especially for
beta-coronaviruses.\textsuperscript{14} To project viral sharing values
into a single value for every pixel, we averaged the pairwise scores.
High values of the average sharing propensity means that this specific
extant bat assemblage is likely to be proficient at exchanging viruses.

@@

To visualize the aggregated risk at the global scale, we combine the
three individual risk components (phylogenetic diversity, compositional
uniqueness, and viral sharing) using an additive color
model.\textsuperscript{\textbf{Seekell2018GeoLak?}} In this approach,
every risk component gets assigned a component in the RGB color model
(phylogenetic diversity is green, compositional uniqueness is red, and
viral sharing is blue). In order to achieve a valid RGB measure, all
components are re-scaled to the {[}0,1{]} interval, so that a pixel with
no sharing, no phylogenetic diversity, and no compositional uniqueness
is black, and a pixel with maximal values for each is white. This
additive model conveys both the intensity of the overall risk, but also
the nature of the risk as colors diverge towards combinations of values
for three risk components. Out of the possible combinations, the most
risky in terms or rapid diversification and spillover potential is high
phylogenetic diversity and low viral
sharing,\textsuperscript{\textbf{Gomulkiewicz2000HotSpo?}} in that this
allows multiple independent host-virus coevolutionary dynamics to take
place in the same location. In the colorimetric space, this correspond
to yellow -- because the HSV space is more amenable to calculations for
feature extraction,\textsuperscript{\textbf{Keke2010StuSki?}} we
measured the risk level by calculating the angular distance of the hue
of each pixel to a reference value of 60 (yellow), and weighted this
risk level by the value component. Specifically, given a pixel with
colorimetric coordinates \((h,s,v)\), its ranged weighted risk value is

@@

To next represent phylogeography of betacoronaviruses in bats, we
aggregated and analyzed betacoronavirus sequence data. We used the
following query to pull all \emph{Betacoronavirus} sequence data from
the GenBank Nucleotide database except SARS-CoV-2;
(``Betacoronavirus''{[}Organism{]} OR betacoronavirus{[}All Fields{]})
NOT (``Severe acute respiratory syndrome coronavirus 2''{[}Organism{]}
OR sars-cov-2{[}All Fields{]}). We added a single representative
sequence for SARS-CoV-2 and manually curated to remove sequences without
the RNA-dependent RNA polymerase (RdRp) sequence or that contained words
indicating recombinant or laboratory strains including ``patent'',
``mutant'', ``GFP'', and ``recombinant''. We filtered over-represented
taxa including betacoronavirus 1, hCoV-OC43, Middle East respiratory
syndrome coronavirus, Murine hepatitis virus, and hCoV-HKU1. Curated
betacoronavirus RdRp sequences were then aligned using
MAFFT\textsuperscript{\textbf{Katoh2013MafMul?}} v1.4.0 (Algorithm
FFT-NS-2, Scoring matrix 200PAM / k=2, gap open penalty 1.53m offset
value 0.123) and a maximum likelihood tree reconstructed in
IQ-TREE\textsuperscript{\textbf{Nguyen2015IqtFas?}} v1.6.12 with
ModelFinder\textsuperscript{\textbf{Kalyaanamoorthy2017ModFas?}}
ultrafast bootstrap
approximation\textsuperscript{\textbf{Hoang2018UfbImp?}} with a general
time reversible model with empirical base frequencies and the
5-discrete-rate-category FreeRaye model of nucleotide substitution
(GTR+F+R5).

We first tested the hypothesis that hotspots of viral diversification
would track hotspots of bat diversification. To do so, we plotted the
number of known bat hosts (specifically only those included in the
phylogeny, so there was a 1:1 correspondence between data sources)
against the ``mean evolutionary distinctiveness'' of the associated
viruses. To calculate this, we derived the fair proportions evolutionary
distinctiveness\textsuperscript{\textbf{Isaac2007MamEdg?}} for each of
the viruses in the tree, then averaged these at the bat species level,
projected these values onto their geographic distributions, and averaged
across every bat found in a given pixel. As such, this can be thought of
as a map of the mean evolutionary distinctiveness of the known viral
community believed to be associated with a particular subset of bats
present.

@@

\leavevmode\hypertarget{ref-Mollentze2020Viral}{}%
\CSLLeftMargin{55. }
\CSLRightInline{Mollentze, N. \& Streicker, D. G. Viral zoonotic risk is
homogenous among taxonomic orders of mammalian and avian reservoir
hosts. \emph{Proceedings of the National Academy of Sciences}
\textbf{117}, 9423--9430 (2020).}

>>>>

We downloaded the data on bats hosts of \emph{Betacoronavirus} from
\texttt{https://www.viralemergence.org/betacov} on
Apr.~2022,\textsuperscript{21} and filtered it to ``known'' hosts
(established before the emergence of SARS-CoV-2) and ``novel'' hosts
(confirmed through sampling and competence assays since the initial data
collection). The original database was assembled by a combination of
data mining and literature surveys, including automated alerts on the
``bats'' and ``coronavirus'' keywords to identify novel empirical
evidence of bats-betacoronaviruses associations; this yielded a total of
126 known hosts, 47 of which were novel hosts. This host--virus list of
interactions was obtained through a comprehensive aggregation of GenBank
data as well as systematic literature searches,\textsuperscript{21,25}
such that we have high confidence in its fitness for the purpose of
inference at a large spatial scale.

@@

We downloaded the rangemap of every current bat species that was
classified as an empirically documented host of \emph{Betacoronavirus}
from the previous step, according to recent IUCN
data.\textsuperscript{\textbf{IUCN2021Iuc?}} The IUCN data have been
assembled to support wildlife conservation efforts, and therefore we do
not expect that they are biased by wildlife disease sampling efforts or
priority. The range maps were subsequently rasterized using the
\texttt{rasterize} function from \texttt{GDAL}\textsuperscript{55} at a
resolution of approximately 100kmx100km at the equator. For every pixel
in the resulting raster where at least one bat host of
\emph{Betacoronavirus} was present, we extract the species pool (list of
all known bat hosts), which was used to calculate the following risk
assessment components: bat phylogenetic diversity, bat compositional
uniqueness, and predicted viral sharing risk.

@@

For every pixel, we measured Faith's Phylogenetic
Diversity\textsuperscript{56} based on a recent synthetic tree with
robust time calibration, covering about 6000 mammalian
species.\textsuperscript{57} Faith's PD measures the sum of unique
branches from an arbitrary root to a set of tips, and comparatively
larger values indicate a more phylogenetic diverse species pool. We
measured phylogenetic diversity starting from the root of the entire
tree (and not from Chiroptera); this bears no consequences on the
resulting values, since all branches leading up to Chiroptera are only
counted once per species pool, and (as we explain when describing the
assembly of the composite risk map), all individual risk components are
ranged in {[}0,1{]}. This measure incorporates a richness component,
which we chose not to correct for; the interpretation of the
phylogenetic diversity is therefore a weighted species richness, that
accounts for phylogenetic over/under-dispersal in some places.

@@

For every species pool, we measured its Local Contribution to
Beta-Diversity;\textsuperscript{58} LCBD works from a species-data
matrix (traditionally noted as \(\mathbf{Y}\)), where species are rows
and sites are columns, and a value of 1 indicates occurrence. We
extracted the Y matrix assuming that every pixel represents a unique
location, and following best practices\textsuperscript{59} transformed
it using Hellinger's distance to account for unequal bat richness at
different pixels. The correction of raw community data is particularly
important for two reasons: first, it prevents the artifact of richer
sites having higher importance; second, it removes the effect of overall
species richness, which is already incorporated in the phylogenetic
diversity component. High values of LCBD indicate that the pixel has a
community that is on average more dissimilar in species composition than
what is expected knowing the entire matrix, i.e.~a more unique
community. Recent results by\textsuperscript{60} shows that LCBD
measures are robust with regards to spatial scale, and are therefore
applicable at the global scale.

@@

For all bat hosts of \emph{Betacoronavirus}, we extracted their
predicted viral sharing network, generated from a previously published
generalized additive mixed model of virus sharing by a tensor function
of phylogenetic distance and geographic range overlap across
mammals.\textsuperscript{61} This network stores pairwise values of
viral community similarity, measured for all hosts (to maintain
consistency with teh phylogenetic diversity measure) across all viruses;
therefore, we consider that it accounts for some overall similarity in
the way hosts deal with viruses, and not only betacoronaviruses. There
is empirical evidence that capacity for cross-species transmission even
between divergent species is generally high,\textsuperscript{62}
especially for beta-coronaviruses.\textsuperscript{14} To project viral
sharing values into a single value for every pixel, we averaged the
pairwise scores. High values of the average sharing propensity means
that this specific extant bat assemblage is likely to be proficient at
exchanging viruses.

@@

To visualize the aggregated risk at the global scale, we combine the
three individual risk components (phylogenetic diversity, compositional
uniqueness, and viral sharing) using an additive color
model.\textsuperscript{63} In this approach, every risk component gets
assigned a component in the RGB color model (phylogenetic diversity is
green, compositional uniqueness is red, and viral sharing is blue). In
order to achieve a valid RGB measure, all components are re-scaled to
the {[}0,1{]} interval, so that a pixel with no sharing, no phylogenetic
diversity, and no compositional uniqueness is black, and a pixel with
maximal values for each is white. This additive model conveys both the
intensity of the overall risk, but also the nature of the risk as colors
diverge towards combinations of values for three risk components. Out of
the possible combinations, the most risky in terms or rapid
diversification and spillover potential is high phylogenetic diversity
and low viral sharing,\textsuperscript{64} in that this allows multiple
independent host-virus coevolutionary dynamics to take place in the same
location. In the colorimetric space, this correspond to yellow --
because the HSV space is more amenable to calculations for feature
extraction,\textsuperscript{65} we measured the risk level by
calculating the angular distance of the hue of each pixel to a reference
value of 60 (yellow), and weighted this risk level by the value
component. Specifically, given a pixel with colorimetric coordinates
\((h,s,v)\), its ranged weighted risk value is

@@

To next represent phylogeography of betacoronaviruses in bats, we
aggregated and analyzed betacoronavirus sequence data. We used the
following query to pull all \emph{Betacoronavirus} sequence data from
the GenBank Nucleotide database except SARS-CoV-2;
(``Betacoronavirus''{[}Organism{]} OR betacoronavirus{[}All Fields{]})
NOT (``Severe acute respiratory syndrome coronavirus 2''{[}Organism{]}
OR sars-cov-2{[}All Fields{]}). We added a single representative
sequence for SARS-CoV-2 and manually curated to remove sequences without
the RNA-dependent RNA polymerase (RdRp) sequence or that contained words
indicating recombinant or laboratory strains including ``patent'',
``mutant'', ``GFP'', and ``recombinant''. We filtered over-represented
taxa including betacoronavirus 1, hCoV-OC43, Middle East respiratory
syndrome coronavirus, Murine hepatitis virus, and hCoV-HKU1. Curated
betacoronavirus RdRp sequences were then aligned using
MAFFT\textsuperscript{66} v1.4.0 (Algorithm FFT-NS-2, Scoring matrix
200PAM / k=2, gap open penalty 1.53m offset value 0.123) and a maximum
likelihood tree reconstructed in
IQ-TREE\textsuperscript{\textbf{Nguyen2015IqTree?}} v1.6.12 with
ModelFinder\textsuperscript{\textbf{Kalyaanamoorthy2017ModelFinder?}}
ultrafast bootstrap approximation\textsuperscript{67} with a general
time reversible model with empirical base frequencies and the
5-discrete-rate-category FreeRaye model of nucleotide substitution
(GTR+F+R5).

We first tested the hypothesis that hotspots of viral diversification
would track hotspots of bat diversification. To do so, we plotted the
number of known bat hosts (specifically only those included in the
phylogeny, so there was a 1:1 correspondence between data sources)
against the ``mean evolutionary distinctiveness'' of the associated
viruses. To calculate this, we derived the fair proportions evolutionary
distinctiveness\textsuperscript{68} for each of the viruses in the tree,
then averaged these at the bat species level, projected these values
onto their geographic distributions, and averaged across every bat found
in a given pixel. As such, this can be thought of as a map of the mean
evolutionary distinctiveness of the known viral community believed to be
associated with a particular subset of bats present.

@@

\leavevmode\hypertarget{ref-RouaultEven2022Gdal}{}%
\CSLLeftMargin{55. }
\CSLRightInline{Rouault, E. \emph{et al.} \emph{GDAL/OGR Geospatial Data
Abstraction software Library}. (Zenodo, 2022).
doi:\href{https://doi.org/10.5281/ZENODO.5884351}{10.5281/ZENODO.5884351}.}

\leavevmode\hypertarget{ref-Faith1992Conservation}{}%
\CSLLeftMargin{56. }
\CSLRightInline{Faith, D. P. Conservation evaluation and phylogenetic
diversity. \emph{Biological Conservation} \textbf{61}, 1--10 (1992).}

\leavevmode\hypertarget{ref-Upham2019Inferring}{}%
\CSLLeftMargin{57. }
\CSLRightInline{Upham, N. S., Esselstyn, J. A. \& Jetz, W. Inferring the
mammal tree: Species-level sets of phylogenies for questions in ecology,
evolution, and conservation. \emph{PLOS Biology} \textbf{17}, e3000494
(2019).}

\leavevmode\hypertarget{ref-Legendre2013Beta}{}%
\CSLLeftMargin{58. }
\CSLRightInline{Legendre, P. \& De Cáceres, M. Beta diversity as the
variance of community data: Dissimilarity coefficients and partitioning.
\emph{Ecology Letters} \textbf{16}, 951--963 (2013).}

\leavevmode\hypertarget{ref-Legendre2019Spatial}{}%
\CSLLeftMargin{59. }
\CSLRightInline{Legendre, P. \& Condit, R. Spatial and temporal analysis
of beta diversity in the Barro Colorado Island forest dynamics plot,
Panama. \emph{Forest Ecosystems} \textbf{6}, 7 (2019).}

\leavevmode\hypertarget{ref-Dansereau2022Evaluating}{}%
\CSLLeftMargin{60. }
\CSLRightInline{Dansereau, G., Legendre, P. \& Poisot, T. Evaluating
ecological uniqueness over broad spatial extents using species
distribution modelling. \emph{Oikos} \textbf{n/a}, e09063 (2022).}

\leavevmode\hypertarget{ref-Albery2020Predicting}{}%
\CSLLeftMargin{61. }
\CSLRightInline{Albery, G. F., Eskew, E. A., Ross, N. \& Olival, K. J.
Predicting the global mammalian viral sharing network using
phylogeography. \emph{Nature Communications} \textbf{11}, 2260 (2020).}

\leavevmode\hypertarget{ref-Mollentze2020Viral}{}%
\CSLLeftMargin{62. }
\CSLRightInline{Mollentze, N. \& Streicker, D. G. Viral zoonotic risk is
homogenous among taxonomic orders of mammalian and avian reservoir
hosts. \emph{Proceedings of the National Academy of Sciences}
\textbf{117}, 9423--9430 (2020).}

\leavevmode\hypertarget{ref-Seekell2018Geography}{}%
\CSLLeftMargin{63. }
\CSLRightInline{Seekell, D. A., Lapierre, J.-F. \& Cheruvelil, K. S. A
geography of lake carbon cycling. \emph{Limnology and Oceanography
Letters} \textbf{3}, 49--56 (2018).}

\leavevmode\hypertarget{ref-Gomulkiewicz2000Hot}{}%
\CSLLeftMargin{64. }
\CSLRightInline{Gomulkiewicz, R., Thompson, J. N., Holt, R. D., Nuismer,
S. L. \& Hochberg, M. E. Hot Spots, Cold Spots, and the Geographic
Mosaic Theory of Coevolution. \emph{The American Naturalist}
\textbf{156}, 156--174 (2000).}

\leavevmode\hypertarget{ref-Keke2010Study}{}%
\CSLLeftMargin{65. }
\CSLRightInline{Keke, S., Peng, Z. \& Guohui, L. Study on skin color
image segmentation used by Fuzzy-c-means arithmetic. in \emph{2010
Seventh International Conference on Fuzzy Systems and Knowledge
Discovery} vol. 2 612--615 (2010).}

\leavevmode\hypertarget{ref-Katoh2013Mafft}{}%
\CSLLeftMargin{66. }
\CSLRightInline{Katoh, K. \& Standley, D. M. MAFFT Multiple Sequence
Alignment Software Version 7: Improvements in Performance and Usability.
\emph{Molecular Biology and Evolution} \textbf{30}, 772--780 (2013).}

\leavevmode\hypertarget{ref-Hoang2018Ufboot2}{}%
\CSLLeftMargin{67. }
\CSLRightInline{Hoang, D. T., Chernomor, O., von Haeseler, A., Minh, B.
Q. \& Vinh, L. S. UFBoot2: Improving the Ultrafast Bootstrap
Approximation. \emph{Molecular Biology and Evolution} \textbf{35},
518--522 (2018).}

\leavevmode\hypertarget{ref-Isaac2007Mammals}{}%
\CSLLeftMargin{68. }
\CSLRightInline{Isaac, N. J. B., Turvey, S. T., Collen, B., Waterman, C.
\& Baillie, J. E. M. Mammals on the EDGE: Conservation Priorities Based
on Threat and Phylogeny. \emph{PLOS ONE} \textbf{2}, e296 (2007).}
